---
title: "Parameter Fitting for Models with Differential Equations"
author: "darthaline"
date: "28 Sept 2020"
output:
  html_notebook:
    toc: true
---

Based on [Learning R: Parameter Fitting for Models Involving Differential Equations](https://notesofdabbler.wordpress.com/2013/06/30/learning-r-parameter-fitting-for-models-involving-differential-equations/) by rdabbler.

[Original code and data](https://bitbucket.org/notesofdabbler/blog_notesofdabbler/src/2932ce48caa370abd549531df5fe4a93bfc26c23/04_learnR_parfitODE/?at=master).

## Loading libraries

```{r}
# load libraries
library(ggplot2) #plotting
library(cowplot) #plotting
library(reshape2) #reshaping data (tall-narrow <-> short-wide)
library(deSolve) #solving differential equations
library(minpack.lm) #least squares fit using levenberg-marquart algorithm
```

## Loading data

```{r digitizedData, eval=FALSE}
data <- data.frame(Time = numeric(0),
                   Conc = numeric(0),
                   Species = character(0),
                   stringsAsFactors = FALSE)

fileList <- list.files(file.path('data'))
fileList <- fileList[grep('.csv', fileList)]

for (i in 1:length(fileList)){
  fileTemp <- read.csv(file.path('data', fileList[i]), header = FALSE)
  fileTemp$Experiment <- sub('.csv', '', fileList[i])
  
  data <- rbind(data, fileTemp,
                stringsAsFactors = FALSE)
}

rm(fileTemp, i, fileList)

colnames(data) <- c('time', 'Conc', 'Species')

# averaging times
time <- data$time
timeList <- c()
for (i in 1:(length(time)/3)) {
  avgVal <- (time[i] + time[i + length(time)/3] + time[i + 2*length(time)/3]) / 3
  timeList <- c(timeList, avgVal)
}

data$time <- rep(timeList, 3)
rm(time, timeList, i, avgVal)
```

```{r}
#load concentration data
df=read.table(file.path('data', "ABC_data.dat"))
names(df)=c("time","ca","cb","cc")
dataMelted=melt(df,id.vars=c("time"),variable.name="species",value.name="conc")

```

## Plotting data

```{r}
plotData <- ggplot(data=dataMelted, aes(x=time, y=conc, color=species))+
  geom_point(size=3)+
  theme_half_open() +
  background_grid()
plotData
rm(plotData)
```

## Equations

$\frac{dc_A}{dt} = -k_1 c_A$

$\frac{dc_B}{dt} = k_1 c_A - k_2 c_B$

$\frac{dc_C}{dt} = k_2 c_B$

with initial concentrations $c_A = 1$, $c_B = 0$, $c_C = 0$.

```{r reactionFunction}
# rate function
rxnrate = function(t, c, parms){
 
 # rate constant passed through a list called parms
 k1 = parms$k1
 k2 = parms$k2
 
 # c is the concentration of species
 
 # derivatives dc/dt are computed below
 r = rep(0, length(c))
 r[1] = -k1*c["A"] #dcA/dt
 r[2] = k1*c["A"] - k2*c["B"] #dcB/dt
 r[3] = k2*c["B"] #dcC/dt
 
 # the computed derivatives are returned as a list
 # order of derivatives needs to be the same as the order of species in c
 return(list(r))
 
}
```

We can compute trajectories for a given parameter set, here $k_1 = 2$, $k_2 = 1$.

```{r}
# predicted concentration for a given parameter set
cinit = c(A = 1,B = 0, C = 0)
t = unique(df$time)
parms = list(k1 = 2, k2 = 1)
out = ode(y = cinit, times = t, func = rxnrate, parms = parms, method = 'euler')
head(out)

rm(cinit, t, parms, out)
```
## Computing residuals

```{r}
ssq = function(parms){
  
  # inital concentration
  cinit = c(A = 1,B = 0,C = 0)
  # time points for which conc is reported
  # include the points where data is available
  t=c(seq(0,5,0.1),df$time)
  t=sort(unique(t))
  # parameters from the parameter estimation routine
  k1 = parms[1]
  k2 = parms[2]
  # solve ODE for a given set of parameters
  out = ode(y = cinit, times = t, func = rxnrate, parms = list(k1 = k1,k2 = k2), method = 'euler')
  
  # Filter data that contains time points where data is available
  outdf = data.frame(out)
  outdf = outdf[outdf$time %in% df$time,]
  # Evaluate predicted vs experimental residual
  preddf=melt(outdf,id.var="time",variable.name="species",value.name="conc")
  expdf=melt(df,id.var="time",variable.name="species",value.name="conc")
  ssqres=preddf$conc-expdf$conc
  
  # return predicted vs experimental residual
  return(ssqres)

}
```

```{r}
# parameter fitting using levenberg marquart algorithm
# initial guess for parameters
parms = c(k1 = 0.5, k2 = 0.5)
# fitting
fitval = nls.lm(par = parms, fn = ssq)

summary(fitval)
```

Estimated parameter values:

```{r estimatedParameters}
parest=as.list(coef(fitval))
parest
```
Number of degrees of freedom (number of datapoints - number of fitted parameters):

```{r degreesOfFreedom}
dof=nrow(dataMelted)-2
dof
```

Mean error:

```{r meanError}
ms=sqrt(deviance(fitval)/dof)
ms
```

Covariance matrix:

```{r covarianceMatrix}
S=vcov(fitval)
S
```

```{r cleanUp}
rm(fitval, ms)
```


## Plotting predicted vs experimental

```{r}
# plot of predicted vs experimental data
 
# simulated predicted profile at estimated parameter values
cinit = c(A = 1, B = 0, C = 0)
t = seq(0,5, 0.2)
parms = as.list(parest)
out = ode(y = cinit, times = t, func = rxnrate, parms = parms)
outdf = data.frame(out)
names(outdf)=c("time","ca predicted","cb predicted","cc predicted")
 
# Overlay predicted profile with experimental data
tmppred=melt(outdf,id.var=c("time"),variable.name="species",value.name="conc")

```


```{r}
library(scales)
hex_codes1 <- hue_pal()(3)

plotData <- ggplot(data=dataMelted, aes(x=time, y=conc, color=species))+
  geom_point(size=3)+
  geom_line(data=tmppred,aes(x=time,y=conc,color=species,linetype=species))+
  theme_half_open() +
  background_grid() +
  scale_color_manual(values=rep(hex_codes1,each=2))
plotData

rm(cinit, t, parms, out, outdf, tmppred, hex_codes1, plotData)
```

## Parameter uncertainty estimation

```{r}
# Get the 95% confidence region
 
# Inverse of covariance matrix
Sinv = solve(S)
 
# get points for a circle with radius r
r = sqrt(qf(0.95, 2, dof)*2)
theta = seq(0, 2*pi, length.out=100)
z = cbind(r*cos(theta), r*sin(theta))
# transform points of circle into points of ellipse using
# svd of inverse covariance matrix
Sinv_svd = svd(Sinv) # inverse of covariance matrix
xt = t(Sinv_svd$v) %*% diag(1/sqrt(Sinv_svd$d)) %*% t(z) # transform from circle to ellipse
x = t(xt)
# translate the ellipse so that center is the estimated parameter value
x = x + matrix(rep(as.numeric(parest),100), nrow = 100, byrow = T)

x <- data.frame(x)
colnames(x) <- c('k1', 'k2')
plotEllipse <- ggplot(x, aes(k1, k2)) +
  geom_point()+
  geom_point(data = data.frame(parest), aes(k1, k2)) +
  theme_half_open() +
  background_grid()
plotEllipse

rm(Sinv, S, r, theta, z, dof, Sinv_svd, xt, x, plotEllipse)
```




